They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-21 paper/paper.Rmd
adding breakdown by round i

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
authors:
  - name: Tony ElHabr
    affiliations:
      - name: Georgia Institute of Technology
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, I find evidence supprting the hypothesis that round win probability can be modeled as a constant across the series, although not at the naive 50%. Second, I examine post-streak round win probability, given the series outcome. I find no evidence that post-streak win rate is significantly different than expected for combinations of streak length, series length, and series winner, suggesting that the perception of momentum at the team level is deceiving.
bibliography: references.bib  
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

```{r}
#| label: setup-code
library(dplyr)
library(qs)
library(tidyr)
library(purrr)
library(gt)
library(scales)

cod_rounds <- qs::qread('data/cod_rounds.qs')
```

@@

## Description of Call of Duty Search and Destroy

@@

The empirical offensive round win percentage across all rounds is $\phi_O$ = 47.8%.[^8] Table \ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from this global average.

@@

@@

cod_round_and_series_win_prop_by_side |>
  filter(is_offense) |>
  transmute(
    pre_cumu_w,
    pre_cumu_l,
    lab = sprintf('%.1f%%\n(%s)', round(100 * win_round_prop, 1), n)
  ) |>
  pivot_wider(
    names_from = pre_cumu_l,
    values_from = lab
  ) |>
  arrange(pre_cumu_w) |>
  gt::gt() |>
  gt::cols_label(
    pre_cumu_w = "Defense round wins",
  ) |>
  gt::tab_spanner(
    label = "Offense round wins",
    columns = `0`:`5`
  ) |>
  gt::as_latex()
```

@@

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^9] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^9]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, refuted the conclusions of GVT, proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^10]. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT.'s data, MS come to the opposite conclusions as GVT.

[^10]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanation for the persistence of gambler's fallacy beliefs", per MS.

@@

Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds). On the other hand, teams in CoD SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^11] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.

[^11]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least one in each league.

@@

The formula for the expected proportion of series a best-of-$s$ format ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^12] is

[^12]: If $\phi > 0.5$, then one might say that this team is the better team (known in hindsight).

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}
```
$s = 11$ rounds for CoD SnD.[^13]

[^13]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to proportions represent probabilities. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while "bare" symbols mirroring hat-ed symbols convey observational data.

@@

As shown by MS with their Theorem 1, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}_{MS}$, to be strictly less than $p^{w|r}$.[^14] Unfortunately, there does not exist a closed form representation for the expectation for $\hat{P}^{w|kr}_{MS}$ for $k > 1$. Nonetheless, one may recursively build a collection of dictionaries to compute the expectation analytically, or, more conveniently, one may run simulations to estimate the expected value.[^15]

[^14]: See Appendix E.1 of MS for details on the proof that follows.

[^15]: I've adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although it would be convenient to emulate the hypothesis testing for difference of proportions performed by MS (and GVT), my context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^16] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{w|kr}$ and $\hat{P}^{\ell|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rates is not completely appropriate, although useful as a reference.

[^16]: GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}
```
using $m$ from Equation \ref{eq:m}. We can perform a binomial test to test the null hypothesis $H_0 : \omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^17]

[^17]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

@@

Stepping back from the one-sided perspective of a single team's probability of winning and their streakiness, I can attempt to detect the "hot hand" more broadly with the Wald-Wolfowitz runs test, a common tool used in the field of quality control. Under the null hypothesis, the number of runs in a sequence of $r$ trials is a random variable with the following mean $\mu$ and variance $\sigma^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r_+r_-}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}
```
# Results

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^18] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^18]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

summarize_cod_win_prop_after_k_wins <- function(k, ...) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    win_series = c(TRUE, FALSE),
    win_round = c(TRUE, FALSE)
  ) |> 
    mutate(
      is_valid = case_when(
        win_series ~ TRUE,
        (n_rounds - !!k - 6L) >= 0L ~ TRUE,
        TRUE ~ FALSE
      )
    ) |> 
    filter(is_valid) |> 
    select(-is_valid) |> 
    mutate(
      naive_prop = ifelse(win_series, 6L / n_rounds, (n_rounds - 6L) / n_rounds),
      notional_prop = (ifelse(win_series, 6L, n_rounds - 6L) - k) / (n_rounds - k),
      across(c(naive_prop, notional_prop), ~ifelse(win_round, .x, 1 - .x)),
      ms_prop = map2_dbl(
        n_rounds, naive_prop, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      ),
      across(
        ms_prop,
        ~ifelse(win_round, .x, naive_prop + (naive_prop - .x))
      ),
      across(
        ms_prop,
        ~ifelse(is.nan(ms_prop), naive_prop, .x)
      )
    )

@@

summarize_cod_win_prop_after_k_wins_i_rounds <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    round = 3L:11L,
    win_series = c(TRUE, FALSE),
    win_round = c(TRUE, FALSE)
  ) |> 
    filter(round >= (!!k + 1L)) |> 
    mutate(
      is_valid = case_when(
        win_series ~ TRUE,
        (n_rounds - !!k - 6L) >= 0L ~ TRUE,
        TRUE ~ FALSE
      )
    ) |> 
    filter(is_valid) |> 
    select(-is_valid) |> 
    mutate(
      naive_prop = ifelse(win_series, 6L / n_rounds, (n_rounds - 6L) / n_rounds),
      notional_prop = (ifelse(win_series, 6L, n_rounds - 6L) - k) / (n_rounds - k),
      across(c(naive_prop, notional_prop), ~ifelse(win_round, .x, 1 - .x)),
      ms_prop = map2_dbl(
        n_rounds, naive_prop, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      ),
      across(
        ms_prop,
        ~ifelse(win_round, .x, naive_prop + (naive_prop - .x))
      )
    )

cod_round_win_prop_after_k_wins <- tibble(win_streak = 2L:5L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_k_wins)
  ) |>
  unnest(data)
cod_round_win_prop_after_k_wins |> filter(win_round, notional_p_value < 0.05)
cod_round_win_prop_after_k_wins |> filter(win_round, ms_p_value < 0.05)

tibble(win_streak = 1L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_k_wins_i_rounds)
  ) |>
  unnest(data) |> 
  filter(win_round, notional_p_value < 0.05)

cod_round_win_prop_after_k_wins_i_rounds <- tibble(win_streak = 1L:3L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_k_wins_i_rounds)
  ) |>
  unnest(data)
cod_round_win_prop_after_k_wins_i_rounds |> 
  filter(notional_prop < prop, notional_p_value < 0.05) |> 
  filter(n_rounds == round) |> 
  filter(win_streak == 1) |> 
  group_by(n_rounds, round) |> 
  mutate(
    total = sum(n),
    prop = n / total
  )
```

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^19] Table \ref{tbl:cod-pw3r-pl3r} compares the observed round win rate $P^{w,kr}$ following streaks of $k=3$ round wins given the series outcome ($r$ rounds and winner) with the notional and MS expected proportions, $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ respectively.

[^19]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

```{=tex}
\begin{longtable}{rrrrrr}
\caption{Observed count and proportion of round wins immediately following round win streak of $k=3$, $N^{w|kr}$ and $P^{w|kr}$ respectively, given the length of the series ($r$ rounds) and the series winner. Additionally, the notional and MS expected proportions, $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ respectively, and the $p$-values of binomial tests comparing the underlying notional and MS round win probabilities, $p^{w|kr}_0$ and $p^{w|kr}_{MS}$, to the probability implied by the observed proportion $p^{w|kr} = N^{w|kr} \times P^{w|kr}$.}
\label{tbl:cod-pw3r-pl3r} \\
\toprule
$r$ & \text{Win series?} & $N^{w|kr}$ & $P^{w|kr}$ & $\hat{P}^{w|kr}_0$ ($p$-value) & $\hat{P}^{w|kr}_{MS}$ ($p$-value) \\ 
\midrule
7 & yes & 209 & 74.6\% & 75.0\% (0.94) & 75.7\% (0.69) \\ 
8 & yes & 209 & 62.2\% & 60.0\% (0.53) & 61.9\% (0.94) \\ 
9 & yes & 193 & 51.8\% & 50.0\% (0.67) & 52.7\% (0.83) \\ 
10 & no & 60 & 13.3\% & 14.3\% (1.00) & 26.7\% (0.02) \\ 
10 & yes & 151 & 43.7\% & 42.9\% (0.87) & 44.7\% (0.87) \\ 
11 & no & 129 & 24.0\% & 25.0\% (0.84) & 30.8\% (0.11) \\ 
11 & yes & 150 & 40.0\% & 37.5\% (0.56) & 38.8\% (0.80) \\

\bottomrule
\end{longtable}
```
We observe that, with the exception of $\hat{P}^{w|kr}_{MS}$ when $r = 10$ and the eventual series loser is the team that wins after a streak of three round wins[^20], all binomial test $p$-values are greater than the typical $\alpha = 0.05$ threshold, implying that we cannot reject the null hypothesis that the expected notional and MS post-streak round win rates are different than the observed proportion. As noted before, the MS expectations may be unreliable for CoD SnD, so one is inclined to prefer the results of the notional binomial tests.

[^20]: This implies that the opponent, th eventual series winner, goes on to win the series 6-4 after going down 0-4. This is the only scenario in which one team wins four rounds consecutively but loses a best-of-11 series in 10 rounds.

@@

# Discussion

For the sake of brevity, I did not delve into win rates and streaks split by offensive or defensive role. For example, it would be interesting to look at whether the round win rate is higher after a streaks of three round wins when the team starts the streak winning as the offensive team. This would mean that, in the round immediately following the streak, the streaking team would be playing defense, where teams are slightly more likely to win on average. On the other hand, streaks of three starting with an offensive round win---theoretically, given $\phi_O$ = 47.7% the expected frequency is $\phi_O^2 \times (1 - \phi_O) \approx 0.119$---are likely to occur less frequently than such streaks starting with a defensive win---$\phi_O \times (1 - \phi_O)^2 \approx 0.130$---so I would want to account for sample size differences.

Regarding future work, I've reached out to the Twitter user "R11stats", who privately tracks in-round player engagements.[^21] R11stats expressed intent on making the data public, which would allow for research into player-specific momentum.

[^21]: <https://twitter.com/R11stats>

>>>>

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
authors:
  - name: Tony ElHabr
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, I find evidence supprting the hypothesis that round win probability can be modeled as a constant across the series, although not at the naive 50%. Second, I examine post-streak round win probability, given the series outcome. I find no evidence that post-streak win rate is significantly different than expected for combinations of streak length, series length, and series winner, suggesting that the perception of momentum at the team level is deceiving.
bibliography: references.bib  
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

```{r}
#| label: setup-code
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(scales)

cod_rounds <- read_csv('data/cod_rounds.csv')
```

@@

## Game Description

@@

The empirical offensive round win percentage across all rounds is $\tau_O$ = 47.8%.[^8] Table \ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from this global average.[^9]

@@

[^9]: At tournaments, higher-seeded ("better") teams choose whether they want to start on offense or defense in either the SnD series played as the second game or the fifth game in the best-of-five format used in tournaments, and their opponent is assigned to play offense in the first round of the SnD slot (second or fifth game) not chosen. [@codrules] (Other game variants like Hardpoint are played as the first, third, and fourth games in the matchup.) Anecdotally, the higher-seeded teams tend to choose to play defense in the first round of the SnD series played as the second game, although this choice is not universal. With online qualifiers, teams all play each other twice throughout the season. Each team gets to play the higher-seeded role once in their matchups, despite their win-loss records.

@@

cod_round_and_series_win_prop_by_side |>
  filter(is_offense) |>
  transmute(
    pre_cumu_w,
    pre_cumu_l,
    lab = sprintf('%.1f%%\n(%s)', round(100 * win_round_prop, 1), n)
  ) |>
  pivot_wider(
    names_from = pre_cumu_l,
    values_from = lab
  ) |>
  arrange(pre_cumu_w) |>
  gt::gt() |>
  gt::cols_label(
    pre_cumu_w = "Defense round wins",
  ) |>
  gt::tab_spanner(
    label = "Offense round wins",
    columns = `0`:`5`
  ) |>
  gt::as_latex() |> 
  cat()
```

@@

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^10] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^10]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, refuted the conclusions of GVT, proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^11]. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT.'s data, MS come to the opposite conclusions as GVT.

[^11]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanation for the persistence of gambler's fallacy beliefs", per MS.

@@

Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds). On the other hand, teams in CoD SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^12] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.

[^12]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least one in each league.

@@

The formula for the expected proportion of series a best-of-$s$ format ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^13] is

[^13]: If $\phi > 0.5$, then one might say that this team is the better team (known in hindsight).

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}
```
$s = 11$ rounds for CoD SnD.[^14]

[^14]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to proportions represent probabilities. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while "bare" symbols mirroring hat-ed symbols convey observational data.

@@

As shown by MS with their Theorem 1, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}_{MS}$, to be strictly less than $p^{w|r}$.[^15] Unfortunately, there does not exist a closed form representation for the expectation for $\hat{P}^{w|kr}_{MS}$ for $k > 1$. Nonetheless, one may recursively build a collection of dictionaries to compute the expectation analytically, or, more conveniently, one may run simulations to estimate the expected value.[^16]

[^15]: See Appendix E.1 of MS for details on the proof that follows.

[^16]: I've adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although it would be convenient to emulate the hypothesis testing for difference of proportions performed by MS (and GVT), my context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^17] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{w|kr}$ and $\hat{P}^{\ell|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rates is not completely appropriate, although useful as a reference.

[^17]: GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}
```
using $m$ from Equation \ref{eq:m}. We can perform a binomial test to test the null hypothesis $H_0 : \omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^18]

[^18]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

@@

Stepping back from the one-sided perspective of a single team's probability of winning and their streakiness, one can attempt to detect the "hot hand" more broadly with the Wald-Wolfowitz runs test, a common tool used in the field of quality control. Under the null hypothesis, the number of runs in a sequence of $r$ trials is a random variable with the following mean $\mu$ and variance $\sigma^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r_+r_-}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}
```
One might subset observed series sequences to those that violate the run test null hypothesis and perform a proportions test where the null is that the observe rate of such series is equal to the frequency that we should expect, given a choice for the constant round win probability $\phi$. If the

# Results

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^19] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^19]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

compute_probs_nx_k <- function(df, k) {
  df |> 
    mutate(
      is_valid = case_when(
        win_series ~ TRUE,
        (n_rounds - !!k - 6L) >= 0L ~ TRUE,
        TRUE ~ FALSE
      )
    ) |> 
    filter(is_valid) |> 
    select(-is_valid) |> 
    mutate(
      naive_prop = ifelse(win_series, 6L / n_rounds, (n_rounds - 6L) / n_rounds),
      notional_prop = (ifelse(win_series, 6L, n_rounds - 6L) - k) / (n_rounds - k),
      across(c(naive_prop, notional_prop), ~ifelse(win_round, .x, 1 - .x)),
      ms_prop = map2_dbl(
        n_rounds, naive_prop, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      ),
      across(
        ms_prop,
        ~ifelse(win_round, .x, naive_prop + (naive_prop - .x))
      ),
      across(
        ms_prop,
        ~ifelse(is.nan(ms_prop), naive_prop, .x)
      )
    )
}

summarize_cod_win_prop_after_k_wins <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    win_series = c(TRUE, FALSE),
    win_round = c(TRUE, FALSE)
  ) |> 
    compute_probs_nx_k(k)

@@

cod_round_win_prop_after_k_wins <- tibble(win_streak = 2L:5L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_k_wins)
  ) |>
  unnest(data)
cod_round_win_prop_after_k_wins |> filter(win_round, notional_p_value < 0.05)
cod_round_win_prop_after_k_wins |> filter(win_round, ms_p_value < 0.05)
# bonferroni correction?
cod_round_win_prop_after_k_wins |> 
  mutate(
    n = n(),
    p2 = notional_p_value |> p.adjust(),
    p3 = ms_p_value |> p.adjust()
  ) |> 
  filter(win_streak == 3, win_round)
```

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^20] Table \ref{tbl:cod-pw3r-pl3r} compares the observed round win rate $P^{w,kr}$ following streaks of $k=3$ round wins given the series outcome ($r$ rounds and winner) with the notional and MS expected proportions, $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ respectively.

[^20]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

```{=tex}
\begin{longtable}{rlrrrr}
\caption{Observed count and proportion of round wins immediately following round win streak of $k=3$, $N^{w|kr}$ and $P^{w|kr}$ respectively, given the length of the series ($r$ rounds) and the series winner. Additionally, the notional and MS expected proportions, $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ respectively, and the $p$-values of binomial tests comparing the underlying notional and MS round win probabilities, $p^{w|kr}_0$ and $p^{w|kr}_{MS}$, to the probability implied by the observed proportion $p^{w|kr} = N^{w|kr} \times P^{w|kr}$.}
\label{tbl:cod-pw3r-pl3r} \\
\toprule
$r$ & \text{Win series?} & $N^{w|kr}$ & $P^{w|kr}$ & $\hat{P}^{w|kr}_0$ ($p$-value) & $\hat{P}^{w|kr}_{MS}$ ($p$-value) \\ 
\midrule
7 & yes & 209 & 74.6\% & 75.0\% (0.94) & 75.7\% (0.69) \\ 
8 & yes & 209 & 62.2\% & 60.0\% (0.53) & 61.9\% (0.94) \\ 
9 & yes & 193 & 51.8\% & 50.0\% (0.67) & 52.7\% (0.83) \\ 
10 & no & 60 & 13.3\% & 14.3\% (1.00) & 26.7\% (0.02) \\ 
10 & yes & 151 & 43.7\% & 42.9\% (0.87) & 44.7\% (0.87) \\ 
11 & no & 129 & 24.0\% & 25.0\% (0.84) & 30.8\% (0.11) \\ 
11 & yes & 150 & 40.0\% & 37.5\% (0.56) & 38.8\% (0.80) \\

\bottomrule
\end{longtable}
```
We observe that, with the exception of $\hat{P}^{w|kr}_{MS}$ when $r = 10$ and the eventual series loser is the team that wins after a streak of three round wins[^21], all binomial test $p$-values are greater than the typical $\alpha = 0.05$ threshold, implying that we cannot reject the null hypothesis that the expected notional and MS post-streak round win rates are different than the observed proportion. As noted before, the MS expectations may be unreliable for CoD SnD, so one is inclined to prefer the results of the notional binomial tests.

[^21]: This implies that the opponent, th eventual series winner, goes on to win the series 6-4 after going down 0-4. This is the only scenario in which one team wins four rounds consecutively but loses a best-of-11 series in 10 rounds.

@@

```{r}
#| label: cod_round_win_prop_after_k_wins_i_rounds
summarize_cod_win_prop_after_k_wins_i_rounds <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    round = 3L:11L,
    win_series = c(TRUE, FALSE),
    win_round = c(TRUE, FALSE)
  ) |> 
    filter(round >= (!!k + 1L)) |> 
    compute_probs_nx_k(k)

cod_round_win_prop_after_k_wins_i_rounds <- tibble(win_streak = 1L:3L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_k_wins_i_rounds)
  ) |>
  unnest(data) |> 
  select(-starts_with('ms')) |> 
  group_by(win_streak, n_rounds, round) |> 
  mutate(
    total = sum(n),
    prop = n / total
  ) |> 
  ungroup() |> 
  mutate(
    notional_p_value = vectorized_binom_test(n, total, p = notional_prop)$p.value |> round(2)
  )

## 5-3?
cod_round_win_prop_after_k_wins_i_rounds |> 
  filter(win_streak == 2) |> 
  filter(n_rounds == 11, round == 11)
```

```{r}
#| label: cod_round_win_prop_after_k_wins_i_rounds-table
# cod_round_win_prop_after_k_wins_i_rounds |> 
#   filter(win_streak == 3) |> 
#   mutate(
#     n = n(),
#     p2 = notional_p_value |> p.adjust()
#   ) |> 
#   filter(!win_round) |> 
#   # filter(notional_p_value < 0.05) |> 
#   filter(n_rounds <= (round + 2)) |> 
#   filter(n_rounds >= 9)
